\LARGE \bf Explicit Knowledge and the Deliberative Layer: Lessons Learned

### 2013-04-23 paper2.tex
Rebased the paper on Severin's IROS2013 paper

@@

\usepackage{graphicx, subfigure}        % standard LaTeX graphics tool
                             % when including figure files
\usepackage{url}
\usepackage{fixme}
\usepackage{lscape}

\journal{Journal of Artificial Intelligence}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Custom macros
\newcommand{\stmt}[1]{{\footnotesize $\langle$\stmttt#1\relax$\rangle$}}
\newcommand{\rawstmt}[1]{{\footnotesize \stmttt#1\relax}}
\def\stmttt#1 #2 #3\relax{{\tt#1} {\bf{\tt #2}} {\tt #3}}

\newcommand{\setstmt}[1]{{\footnotesize [\setstmttt#1\relax]}}
\def\setstmttt#1,#2\relax{\rawstmt{#1}, \rawstmt{#2}}

\newcommand{\concept}[1]{{\footnotesize \texttt{#1}}}

\newcommand{\ie}{{\textit{i.e.~}}}
\newcommand{\cf}{{\textit{cf~}}}
\newcommand{\eg}{{\textit{e.g.~}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\begin{frontmatter}

\title{When the robot considers the human... }
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{Rachid Alami, Mathieu Warnier, Julien Guitton, S\'{e}verin Lemaignan and Emrah Akin Sisbot}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\address{ LAAS-CNRS\\ 
        Universit\'e de Toulouse, UPS, INSA, INP, ISAE, LAAS\\ 
         F-31077 Toulouse, France\\ \texttt{Firstname.Name@laas.fr}}

\begin{abstract}
  This paper addresses some key decisional issues that are
  necessary for a cognitive robot which shares space and tasks with a
  human. We adopt a constructive approach based on the identification
  and the effective implementation of individual and collaborative
  skills. The system is comprehensive since it aims at dealing with a
  complete set of abilities articulated so that the robot controller
  is effectively able to conduct in a flexible manner a collaborative
  task with a human partner. These abilities include geometric
  reasoning and situation assessment based essentially on
  perspective-taking and affordances, management and exploitation of
  each agent (human and robot) knowledge in a separate cognitive
  model, human-aware task planning and human and robot interleaved
  plan achievement.
\end{abstract}

@@

\section{Introduction}

@@

\section{A typical human-robot interaction problem}\label{sec:problem}

@@

\section{Related work and vision}\label{sec:soa}

@@

\section{The LAAS deliberative layer}
\label{sec:Framework}

@@

\begin{figure*}
        \centering
        \includegraphics[width=1.0\columnwidth]{figs/archi}
        \caption{Overview of the LAAS deliberative layer. Knowledge is
        centrally managed in an active \emph{semantic blackboard}, pictured
        above with a thick border.}
        \label{fig|archi}
\end{figure*}

This architecture moves away from standard layered approaches. Interactions
between components are mostly bidirectional and, from the software components
point of view, we do not introduce layers of abstraction (we do, however, have
access to the lower level modules of the robot to execute
actions~\cite{Alami1998}, but all cognition-related modules reside at the same
level). This is especially visible for the dialogue input processing. This
component does not simply act as an alternative perceptual input to the
symbolic database, but also actively queries previously acquired knowledge to
disambiguate and validate the newly created symbolic knowledge (see
section~\ref{sect|com}).

@@

\subsection*{Knowledge model}

@@

\subsection*{Interaction model}

Interaction happens as a consequence of an explicit request of the
human to satisfy a goal or because the robot finds itself in a
situation where it is useful if not mandatory. In both cases, the
robot has a goal to satisfy.  An important issue is the notion of
engagement, a process in which the robot will have to establish,
maintain and terminate a connection with a human partner. 
This covers goal establishment, selection of an incremental refinement
of the task that is intended to be achieved, and execution control
including monitoring, and even influencing, human task performance and
his/her commitment to the goal. The human involvement may range from a
direct participation to the task achievement, to a simple
``acceptance'' of robot activity in his/her close vicinity.

Next sections describe how the robot is controlled through an analysis of the
three main activities performed by the robot controller:

\begin {enumerate}
\item Situation assessment and context management 
\item Goals and plans management
\item Action refinement, execution and monitoring
\end {enumerate}

These activities are caried out by different software composants:

    \item Situation assessment is managed by \textsc{Spark} (Spatial Reasoning
        and Knowledge module)~\cite{Sisbot2011},

    \item HATP (Human-Aware Task Planner)~\cite{Alili2008} is responsible for
        symbolic task planning,

    \item We finally also rely on a set of human-aware motion, placement and
        manipulation planners, grouped under the \textsc{Move3D}
        umbrella~\cite{Sisbot2008, Mainprice2011, Pandey2010}.

Other decisional activities, such as situated dialog (\cite{Ros2010b,
  Lemaignan2011}, not presented here) have been developed that use the
same set of components.

\section{Situation assessment and context management}\label{sec:situ}

This activity involves the geometric and temporal reasoning component,
the symbolic facts and belief management component and the dedicated
robot controller activity (Figure~\ref{architecture_fg}).

We assume that perception provides in real-time the identity and the
position of objects when they are in the field of view of the sensors.
In our implemented examples, the robot is localised using a standard
horizontal laser-scanning based localisation system, the objects are
identified and localized using ARToolkit \cite{ARToolkit} and the
humans are tracked using a commercial motion capture system and a
Kinect device from Microsoft.

\subsection{Geometric and Temporal Reasoning component}\label{sub:gtrc}

The geometric reasoning component plays a central role in our
architecture. It is called SPARK (Spatial Reasoning and
Knowledge~\cite{Sisbot2011}) in the current implementation. It is
responsible for geometric information gathering and it embeds a number
of decisional activities linked to abstraction (symbolic facts
production) and inference based on geometric and temporal reasoning.
SPARK maintains all geometric positions and configurations of agents,
objects and furniture coming from perception and previous or {\it a
  priori} knowledge.

 %\subsubsection*{Symbolic facts production:} 
\vspace{0.3cm}
\noindent
\textbf{Symbolic facts production:} 
Geometric state of the world is abstracted in symbolic facts that can
be classified in three different categories.

@@

%\newline\newline
%\subsubsection*{Hypotheses on objects states and positions}
\vspace{0.3cm}
\noindent
\textbf{Hypotheses on objects states and positions:}
It is sometimes difficult or even impossible to see and/or track an
object in certain states. This happens, for instance, when the object
has been put in a container, when it is in the robot gripper or in the
human hand, and more generally in any state in which it is hidden by
something else. Our robot has a model of the possible symbolic states
for an object (whether the object is on a furniture, in an agent hand,
in a container, etc.).  According to the robot perception of what has
happened since the object was last seen, the robot tries to maintain a
belief of the current possible symbolic states and their associated
probabilities for this object. Such information can be used to update
the beliefs using input from exploration, dialog, human visual
focus,\ldots
SPARK currently provides a simple implementation of such a functionality. The
only managed hypotheses are \emph{in container} and \emph{in agent hand}. We
can have only one hypothesis at the same time. Hypothesis validity is checked
geometrically in case of incoming perception values.

%\subsubsection*{Primitive action recognition}
\vspace{0.3cm}
\noindent {\bf Primitive action recognition:} 
Monitoring human activity is crucial to maintain a coherent state of
the world. Full human action and activity monitoring is a difficult
task that requires knowledge and reasoning both on high level facts
like goals, intentions and plans, as well as bottom-up data from agent
and object motions. Simple temporal and geometric reasoning on human
hand trajectories and potential objects placements can provide some
useful clues for high level human monitoring processes. We call this
temporal and geometric reasoning \emph{primitive action recognition}.

@@

@@

ORO also provides an event mechanism that allows components to be
triggered when specific events occur. A component can
for instance subscribe to events of kind \setstmt{?agent isVisible
  true, ?agent type Human}. As soon as the perception layer detects a
human in the robot's field of view and accordingly updates the
knowledge base, the executive layer would be triggered back. The
event framework also takes advantage of the inference capabilities of
ORO. Thus an event can be indirectly triggered if its triggering
conditions can be inferred to be true.

@@

\section{Goal and Plan Management}\label{sec:plan}

The Goal and Plan Management activity involves the human-aware
symbolic task planner component and the dedicated robot controller
activity (Figure~\ref{architecture_fg}).

\subsection{Symbolic Task Planning}

@@

\subsection{ Goal and plan Controller}
Figure~\ref{goal_plans_fg} sums up the Goal and Plan Management as
implemented in the robot controller.  When an event announcing a new
goal is caught by the controller, the validity of this goal is tested:
does it corresponds to capabilities of agents? is it not already
achieved? Then, the goal is sent to HATP which produces a first plan.
A goal is considered achievable as long as the planner computes a
valid plan and it is not abandoned by the human.

Plan execution consists in the management of all the actions of the
plan. Human and robot are not acting both at the same time.  In case
of plan failure a new plan is requested and executed.

The management at the action level is done in three steps. First, the
action preconditions are tested over the current state of the
world. Then the action is executed and monitored (only monitored of it
corresponds to a human action).  Finally, the expected effects are
verified in order to acknowledge the action achievement.

Concerning the speech acts during plan management, the robot informs
its partners on the goal existence and status, plan existence and
status, ongoing plan action, plan action failure, failing facts for
action precondition or effect assessment.

\begin{figure}[thpb]
  \centering
  \includegraphics[width=0.8\columnwidth]{./figs/plan_management.pdf}
  \caption {Automaton for goal and plan management}
  \label{goal_plans_fg}
\end{figure}

\section{Action execution and monitoring}\label{sec:action}

%\subsubsection*{Human aware motion, placement and manipulation planning}
\vspace{0.3cm}
\noindent
\textbf{Human-aware motion, placement and manipulation planning:}
In our scenarios, actions are only object manipulation actions: Pick,
Place, Throw, Give. Motion and Task planning allows to compute final
object placement, grasp and arm motion trajectories taking into
account task specific constraints and human postures, abilities and
preferences: see \cite{Sisbot2008, Mainprice2011, Pandey2010} for
details.

%\subsubsection*{Action Execution and Monitoring Robot Controller}
\vspace{0.3cm}
\noindent
\textbf{Execution and Monitoring Robot Controller:} Depending on the
context and on the shared plan elaborated by HATP for a given goal,
the robot controller decides to execute an action or to ask its human
partner to do it.  Actions feasibility by the human or the robot are
regularly reconsidered based on the reachability / visibility
computation mechanisms.

\section{An illustrative example}\label{sec:expes}

@@

\section{Conclusion and future work}

@@

The design choices and the results presented here are still preliminary.
While the general scheme we propose might be difficult to implement in
a general sense, we believe that it is a reasonable challenge to
implement it in the case of a personal robot assistant essentially
devoted to fetch-and-carry, as well as for interactive manipulation
tasks and associated activities.\\

\vspace{0.3cm}
\noindent
{\bf Acknowledgements:}
This work has been conducted within the EU CHRIS project ({\tt
  \url{http://www.chrisfp7.eu/}}) funded by the E.C. Division FP7-IST
under Contract 215805. We also would like to thank warmly a number of LAAS
robotics team members and more particularly Xavier Broquere, Mokhtar
Gharbi, Wuwei He, Matthieu Herrb, Jim Mainprice, Raquel Ros, Daniel
Sidobre, Thierry Sim\'{e}on, Muhammad Ali.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8

@@

\usepackage{cite}
\usepackage{graphicx, subfigure}
%\usepackage{hyperref}
\usepackage{url}
\graphicspath{{figs/}}
\DeclareGraphicsExtensions{.pdf,.jpg,.png}

\usepackage{ifthen}

\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

\usepackage[draft]{fixme}

\usepackage{listings}
\usepackage{alltt}
\usepackage{fancyvrb}
\usepackage{flushend} % equalize the column length on last page

\newcommand{\concept}[1]{{\small \texttt{#1}}}

\newcommand{\stmt}[1]{{\footnotesize \tt $\langle$ #1\relax$\rangle$}}
%\newcommand{\stmt}[1]{{\footnotesize $\langle\stmttt#1\relax\rangle$}}
%\newcommand{\rawstmt}[1]{{\footnotesize \stmttt#1\relax}}
%\def\stmttt#1 #2 #3\relax{
%\texttt{\IfBeginWith{#1}{?}{\textbf{#1}}{#1} \IfBeginWith{#2}{?}{\textbf{#2}}{\emph{#2}} \ifthenelse {\equal{#3}{true} \OR \equal{#3}{false}}{\emph{#3}}{\IfBeginWith{#3}{?}{\textbf{#3}} {#3}}}}

%\newcommand{\setstmt}[1]{{\footnotesize [\setstmttt#1\relax]}}
%\def\setstmttt#1,#2\relax{\stmttt#1\relax, \stmttt#2\relax}

\newcommand{\ie}{{\textit{i.e.\ }}}
\newcommand{\cf}{{\textit{cf\ }}}
\newcommand{\eg}{{\textit{e.g.\ }}}

\begin{frontmatter}

\title{\LARGE \bf
Explicit Knowledge and the Deliberative Layer: Lessons Learned
}

\author{Séverin Lemaignan and Rachid Alami}

\address{
CNRS, LAAS, 7 avenue du Colonel Roche, F-31400 Toulouse, France\\
Univ de Toulouse, LAAS, F-31400 Toulouse, France\\
{\tt severin.lemaignan@laas.fr}, {\tt rachid.alami@laas.fr}
}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}

This paper addresses some key decisional issues that are necessary for a
cognitive robot which shares space and tasks with a human. We adopt a
constructive approach based on the identification and the effective
implementation of individual and collaborative skills. The system is
comprehensive since it aims at dealing with a complete set of abilities
articulated so that the robot controller is effectively able to conduct in a
flexible manner a collaborative task with a human partner. These abilities
include geometric reasoning and situation assessment based essentially on
perspective-taking and affordances, management and exploitation of each agent
(human and robot) knowledge in a separate cognitive model, human-aware task
planning and human and robot interleaved plan achievement.

This article presents our design choices, the articulations between the diverse
deliberative components of the robot, and the strengths and weaknesses of our
approach. We show that explicit knowledge management is not only a convenient
tool from the software engineering point of view, but also pushes for a
different, more \emph{semantic} way to address the decision-making issue in
human-robot interactions.

\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The challenge of human-robot interaction}

@@

\subsection{The typical human-robot interaction problem}\label{sec:problem}

@@

\subsection{Towards the cognitive robot at LAAS}

Natural interaction and cooperation are on the (dare we say,
\emph{short-term}) agenda for the human-robot interaction community. They are
keys to the broad class of \emph{interactive manipulation problems}: several
agents agree on a (more or less implicit) joint goal that requires some sort of
cooperation to be successfully achieved. This class of problems involves both
dialogue and manipulation and they are often initially underspecified: they
require iterative and interactive resolution.

Over the last years we have focused our efforts on identifying the cognitive
prerequisites of these challenges, and giving them experimental reality on the
robots: what is required for sentences like ``Let's dress the table together''
to be understood by the robot, correctly interpreted in the spatial and
temporal context of the interaction, and eventually transformed into a set of
actions.

We have chosen to tackle the challenge from several ends: human-aware
navigation and motion planning (\cite{Mainprice2011}), situation assessment
coupled with motion planning (\cite{Mainprice2012}), projection of
``mightabilities'' that anticipates what surrounding agents may do
(\cite{Pandey2011}), and the design and deployment of a pervasive
\emph{knowledge-centric} software architecture.

The last item is the focus of this paper: what does ``knowledge-centric''
architecture means, and what are its benefits and drawbacks.

\subsection{Related work}\label{sec:soa}

@@

\subsection{Article overview}

In the next section, we first introduce the deliberative layer of our robots at
the functional level. It represents what we call a \emph{knowledge-oriented}
architecture, built around a central knowledge management tool called the {\sc
Oro} server.

The following sections go over each of the facets of this deliberative layer:
how knowledge is produced asynchronously from geometric reasoning, how the
robot conducts grounded multi-modal interaction with humans, how knowledge is
used by decision-making components like the robot controller or the task
planner, and finally, how our knowledge-centric architecture enables the
implementation of specific internal cognitive processes.

We conclude the article with a discussion about the strengths and weaknesses of
such a knowledge-oriented architecture.

\section{The LAAS deliberative layer}

@@

\begin{figure*}
        \centering
        \includegraphics[width=1.0\columnwidth]{archi}
        \caption{Overview of the LAAS deliberative layer. Knowledge is
        centrally managed in an active \emph{semantic blackboard}, pictured
        above with a thick border.}
        \label{fig|archi}
\end{figure*}

This architecture moves away from standard layered approaches. Interactions
between components are mostly bidirectional and, from the software components
point of view, we do not introduce layers of abstraction (we do, however, have
access to the lower level modules of the robot to execute actions, but all
cognition-related modules reside at the same level). This is especially visible
for the dialogue input processing. This component does not simply act as an
alternative perceptual input to the symbolic database, but also actively
queries previously acquired knowledge to disambiguate and validate the newly
created symbolic knowledge (see section~\ref{sect|com}).

@@

\subsection{Knowledge model}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Situation Assessment and Context Management}
\label{sect|sit-ass}

\begin{figure}
        \centering
        \includegraphics[width=\columnwidth]{spark-oro}
    \caption{Functional overview of knowledge base ({\sc Oro} server, top part) and the geometric situation assessment module (\emph{SPARK}, bottom part)}
        \label{fig|spark-oro}
\end{figure}

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module (as named on Fig.~\ref{fig|archi}) called SPARK (\emph{SPAtial
Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a situation assessment
reasoner that generates symbolic knowledge from the geometry of the
environment with respect to relations between objects, robots and humans,
also taking into account the different perspective that each agent has on
the environment (Fig.~\ref{fig|spark-oro}).

SPARK is an \emph{amodal} geometric model of the environment that serves both
as basis for the fusion of the perception modalities and as bridge with the
symbolic layer. This geometric model is continuously updated at run-time by the
robot based on its sensors. In our experiments, objects are identified and
localised through 2D barcodes.  On the other hand, the human pose is tracked
with a Kinect-like device (assisted by motion capture to accurately track the
head motion, which is required to compute what the human is looking at).

\subsection{Situation Assessment}

\subsubsection*{Symbolic facts production:}

Geometric state of the world is abstracted in symbolic facts that can
be classified in three different categories.

@@

\subsubsection*{Hypotheses on objects states and positions}
It is sometimes difficult or even impossible to see and/or track an
object in certain states. This happens, for instance, when the object
has been put in a container, when it is in the robot gripper or in the
human hand, and more generally in any state in which it is hidden by
something else. Our robot has a model of the possible symbolic states
for an object (whether the object is on a furniture, in an agent hand,
in a container, etc.).  According to the robot perception of what has
happened since the object was last seen, the robot tries to maintain a
belief of the current possible symbolic states and their associated
probabilities for this object. Such information can be used to update
the beliefs using input from exploration, dialog, human visual
focus,\ldots
SPARK currently provides a simple implementation of such a functionality. The
only managed hypotheses are \emph{in container} and \emph{in agent hand}. We
can have only one hypothesis at the same time. Hypothesis validity is checked
geometrically in case of incoming perception values.

\subsubsection*{Primitive action recognition}
Monitoring human activity is crucial to maintain a coherent state of
the world. Full human action and activity monitoring is a difficult
task that requires knowledge and reasoning both on high level facts
like goals, intentions and plans, as well as bottom-up data from agent
and object motions. Simple temporal and geometric reasoning on human
hand trajectories and potential objects placements can provide some
useful clues for high level human monitoring processes. We call this
temporal and geometric reasoning \emph{primitive action recognition}.

@@

\subsection{Ego-centric and allo-centric frames}

SPARK enables \emph{perspective-taking}: spatial relations between entities can
be computed from different viewpoints, which let the robot build for each
agents it interacts with a different, \emph{perspective-aware} symbolic model
of the environment. These models are separately stored in the knowledge base.

This allows us to deal with ambiguities that arises when one speaker refers to
an object within a reference system (or changes the reference system, \ie
switches perspective) without making the reference frame
explicit~\cite{Breazeal2006, Ros2010}.

As a result, the robot stores models on the environment either in the
\emph{ego-centric} reference frame (from the robot perspective) or in
addressee-centred (\ie, \emph{allo-centric}) frames.

@@

ORO also provides an event mechanism that allows components to be
triggered when specific events occur. A component can
for instance subscribe to events of kind \stmt{?agent isVisible
  true, ?agent type Human}. As soon as the perception layer detects a
human in the robot's field of view and accordingly updates the
knowledge base, the executive layer would be triggered back. The
event framework also takes advantage of the inference capabilities of
ORO. Thus an event can be indirectly triggered if its triggering
conditions can be inferred to be true.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Communication}
\label{sect|com}

\subsection{Natural language grounding}

Natural language processing is one of the fields of human-robot interaction
where the introduction of the semantic layer has been most beneficial.
We detail in~\cite{Lemaignan2011a} the techniques and the tool called
\texttt{dialogs} that we have developed for English language parsing and
grounding, along with verbalisation and (minimalist) dialogue management.

Natural language input (in experiments, we rely on an Android-based interface,
with Google speech recognition) is parsed into a grammatical structure, and
each sentence atoms are resolved with the help of the ontology to ground
concepts like objects (\ie when a user says ``pick the can'', resolve to which
instance of \emph{can} the user is referring to) and actions. Sentences are
sorted into questions, desires and statements, and processed accordingly.

The system supports quantification (``give me \{a | the | some | all | any | ...\} can''),
thematic roles (action-specific predicates that qualify the actions), interactive disambiguation (the robot asks
questions when it needs more information), anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also permits knowledge extension by learning new semantic
structures (for instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal}, interprets common temporal and place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm tired'') into
\emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1 hasFeature
tired}).

\subsection{Limits of disambiguation at semantic level}

One prototypical example of semantic disambiguation has been given in
\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
each other with a set of random objects in-between, one player mentally choose
one object, and the other player has to guess the object by asking closed
questions like \emph{Is your object small or large?} Based on the knowledge it
has acquired, the robot is able to minimize the number of questions required to
find the object.

When playing this kind of game, however, the issue arises that the robot has no
way to select which knowledge about the object is relevant in the interaction
context. For instance, the knowledge base may store facts like \stmt{obj1 type
ActiveConcept} (which internally means that this concept was mentioned in a
discussion in the last few seconds): this information is not a relevant
property of \concept{obj1} when trying to disambiguate concepts with humans.
This distinction between \emph{internal knowledge} (meaningful to
the system only) and \emph{common knowledge} (whose meaning is understood by
all the interactors) has not been properly dealt with in our architecture.

Besides, even knowledge that belongs the \emph{common knowledge} may not be
appropriate in a given interaction context. For instance, the system may
compute that at a given instant the human is looking at the object: \stmt{human
looksAt obj1}. This property makes sense to both parties, but in the context of
the \emph{spygame}, we would like to mainly use immanent properties, not
volatile like a gaze. More research is required to identify relevant
interaction contexts and knowledge classes attached to them.

\subsection{Multi-modal communication}

Because all components rely on the same RDF formalism to format their outputs,
the different communication modalities (\emph{explicit} like verbal, deictic or
based on gaze, or \emph{implicit} like postures) are presented in an
homogeneous way.

The dialogue grounding process makes use of them at two distinct levels.

First, particular steps of the grounding process explicitly check for the
presence and value of specific facts: for instance, when a set of instances
match a category (the human says ``give me the bottle'' and the robot knows
about three bottles), the module may decide (it actually depends on the
quantifier preceding the class) to discard some of them based on their
\emph{visibility} by the speaker (implicit communication context built on the
human posture).

Another example, when the human says ``this'', the robot checks if the human is
currently pointing at some object. In that case, \emph{this} is replaced by the
object focused on.

Note that, while the system benefits from the complementary modalities, they
are not all required. The dialogue system would for instance run equally well
with only the verbal modality.

The second level of integration of multi-modality is implicit: by computing
symbolic properties from the geometry, richer descriptions and hence
discrimination possibilities are available: for instance, if reachability is
available, the robot may asks ``do you mean the bottle that is accessible to
me?'' to discriminate between the three bottles. That way, procedures relying
on discrimination transparently benefit from added modalities.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Robot control}
\label{sect|ctrl}

\subsection{Desires and experiences}

Our robot execution controllers (OpenPRS-based {\sc Shary} or Python-based {\sc
pyRobots}) have deep integration with the knowledge base. It serves as the primary source of information for semantic-aware decision-making.

We split the interaction situations stemming from the situation assessment and
communication components in two categories: \emph{desires} (performative act)
and \emph{experiences} (assertive act).

\emph{Desires} are typically human orders (``Give me that book''). The nature
of the desired action (to pick, give, look, bring, show...), along with the
action parametrization (what is acted on? who should perform the action? etc.)
are extracted from the knowledge base, and either passed to a task planner
(presented in the next section) or executed if the procedure is directly
available.

\emph{Experiences}, on the other hand, comprise of emotions, states and
questions (when asking a question, we consider the human to be in an
\emph{interrogative state}). When the knowledge base recognizes that an agent
\emph{experiences} a particular emotion or state, the execution controller may
decide to handle it, typically by trying to answer the question or using the
emotional or physical state as a parameter for subsequent actions. As an example,
when the speaker says ``I feel tired'', we change the motion planner
parametrization to lower the effort the human need to provide for the following
joint manipulation tasks. Note that this example has been implemented as a
proof-of-concept. We have not yet tried to define a theoretical framework that
would support action alteration based on the user's experienced states.

\subsection{Event-driven control}

The {\sc Oro} server proposes two paradigms to access its content: RPC-style
queries (based on the standard SPARQL language) or events. In its simplest
form, a module can register an event by passing through an event pattern, which
is basically a partial statement like \stmt{? type Elephant}, and a callback.
Every time a new instance of elephant appears in the knowledge base, the
callback is triggered.

This allows us to write reactive robot controllers with a high level of
expressiveness: for instance, by subscribing to the event \stmt{human1 desires
?action, ?action type Look, ?action hasGoal myself}, we could trigger a
behaviour when the human expresses (through dialogue, gestures...) that he
wants to look at the robot itself.

The robot controller designer does not need to directly care about how this
\emph{desire} is produced (this is delegated to perception modules), it can
focuses on the semantic of the desire.

Note also that we take advantage of the reasoning capabilities of the system:
for example, the goal of the action (\stmt{action hasGoal myself}) may well not
have been explicitly asserted, but inferred by the reasoner based on other
assertions.

\subsection{Action execution and monitoring}\label{sec:action}

\subsubsection*{Human aware motion, placement and manipulation planning}
In our scenarios, actions are only object manipulation actions: Pick,
Place, Throw, Give. Motion and Task planning allows to compute final
object placement, grasp and arm motion trajectories taking into
account task specific constraints and human postures, abilities and
preferences: see \cite{Sisbot2008, Mainprice2011, Pandey2010} for
details.

\subsubsection*{Action Execution and Monitoring Robot Controller}
context and on the shared plan elaborated by HATP for a given goal,
the robot controller decides to execute an action or to ask its human
partner to do it.  Actions feasibility by the human or the robot are
regularly reconsidered based on the reachability / visibility
computation mechanisms.

\section{Human-aware task planning}

Our execution controllers rely on symbolic task planning to convert faraway
desires into a succession of atomic actions. We use in our architecture the
HATP planner (for Human Aware Task Planner~\cite{Alili2009}).  HATP is based on
a Hierarchical Task Network (HTN) refinement, which performs an iterative task
decomposition into sub-tasks until reaching atomic actions.  The \emph{planning
domain} defines the set of methods describing how to decompose a task and can
be seen as the procedural knowledge of the robot.

@@

Two important remarks: because HATP is a generic symbolic task planner, we have
been able to design a planning domain at a semantic level which is close to the
one used in the human-robot dialogue (the planner vocabulary contains concepts
like \texttt{give}, \texttt{table}, \texttt{is on}...). Hence only a few
ontology rules have been required to map both the knowledge extracted from the
situation assessment and the statements originated from the verbal interaction
to the planner domain.

Second remark, after some research (see appendix B of~\cite{Lemaignan2012a} for
a detailed discussion)  we have decided not to represent neither the planning
domain nor the resulting plans in the knowledge base: the planning domain (with
tasks pre- and postconditions) is stored in a specific format, outside of the
knowledge repository, and the plans are directly communicated to the robot
controller. From our experience, the OWL formalism is not a satisfying
representation language for concepts that are both tightly bound to time
representation and non-monotonic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Internal cognitive processes}
\label{sect|intern}

\subsection{Theory of Mind}
\label{sect|tom}

Theory of Mind (originally defined in~\cite{Premack1978}) is the cognitive
ability that a subject possesses to represent the mental state of another
agent, possibly including knowledge that contradicts the subject own model: for
example, a book can be at the same time \emph{visible} by myself, and \emph{not
visible} by you.

This skill, which is essential to understand others' perspectives during
interactions, appears in children around their third year. It supposes the ability to build, store and retrieve separate models of the knowledge of the interactors.

Our knowledge base implements such a mechanism: when the robots infers a new
agent has been introduced in the knowledge base, it initializes a new,
independent, ontology for this agent. All the ontologies that are created share
the same common-sense knowledge, but rely on each agent's perspective for the
actual instantiation: the robot (geometrically) computes that the book is in
its own field of view, but not in the human one. The robot knowledge contains
the fact \stmt{book isVisible true} while the human model contains \stmt{book
isVisible false}.

One classical application of this cognitive skills is the so-called
\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
experiment)~\cite{Leslie2000}: a child is asked to watch a scene where two
people, A and B, manipulate objects. Then A leaves and B hides away one
object. When A comes back, we ask the child ``where do you think A will
look for the object?''. Before acquiring a theory of mind, children are not
able to separate their own (true) model of the world (where they know that
the object was hidden) from the model of A, which contains \emph{false
beliefs} on the world (A still thinks the object is at its original
position since he did not see B hiding it). Using seperate knowledge models
in the knowledge base, we have been able to replicate this experience on
our robots~\cite{Warnier2012a}.

\subsection{Working memory}

The {\sc Oro} server also feature a mechanism to mimic simple forms of
biological memory.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.

Three such profiles are predefined: {\tt short term}, {\tt episodic} and {\tt
long term}. They each correspond to a different lifetime for the statements
(respectively 10 seconds, 5 minutes and no time limit). After this duration,
the statements are automatically removed from the knowledge base.

This approach is limited. In particular, \emph{episodic} memory primarily
refers to the semantics of the statements (that is expected to be related to an
event) and not to a specific life duration.

We rely however on this short term memory for a particular use-case:
\emph{active concept}. Some modules, like the natural language processor, use
the {\tt short term} memory profile to mark for a few seconds important
concepts that are currently manipulated by the robot. For example, if a human
asks the robot: ``Give me all red objects'', the human, the \concept{Give}
action, and each red objects that are found are successively marked as
\emph{active concepts} by inserting statements like \stmt{human type
ActiveConcept} in the short-term memory (which can be considered, in this case,
to be a working memory). We use this feature to give a (visual) feedback to the
users.

\section{Experimental results}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion and conclusion}
\label{sect|conclusion}

\subsection{Knowledge-oriented architecture}

Altogether, the components we have presented compose an architecture that we
call \emph{knowledge-oriented}:

    \item{Knowledge is explicitly stored in one central and consistent
    repository of facts, accessible by all modules.}

    \item{Knowledge is represented in a strict formalism (OWL statements) and
    with a clearly defined vocabulary (stated in the common-sense ontology).}

    \item{The first two points enable both a loosely-coupled architecture where
    modules can very easily be removed or replaced by other ones as long as
    they share the same semantics (modules are defined by the knowledge they
    produce),}

    \item{and a \emph{symbolic}, reactive, event-driven approach to robot control.
    By managing events at the same level as the reasoner, we take full
    advantage of the inference abilities of {\sc Oro} to trigger events whose
    \texttt{true} conditions can be inferred.}

    \item{Finally, this architecture allows for the combination of very
    different knowledge modalities in a single homogeneous environment,
    bringing mutual benefits to components. For instance, the dialogue
    processing module can perfectly run without any geometric
    perception, but its disambiguation routines can transparently
    benefit from it when available (since richer symbolic descriptions of
    objects are then available).}

We believe that this \emph{knowledge-oriented} approach has a strong potential
not only to enable rich human-robot interaction, but also as a broader approach
to information alignment and fusion in complex robotic systems.  The
versatility of this paradigm could be illustrated by a simple imaginary
scenario with a blind robot and a deaf robot. The blind robot does not see (no
cameras or alike), but someone can verbally describe a scene to it. On the
other hand, the deaf robot has a good vision system, but cannot process verbal
input.  Without any changes to the software architecture that we described,
control modules of both robots could equally perform the same tasks since all
the knowledge is abstracted and centralised (note that to actually implement
this imaginary situation, the blind robot would of course need \textit{a
priori} 3D models of objects talked about to enable planning or pick and place
actions, and the deaf robot would require at least some gesture interpretation
to understand orders).

\subsection{Conclusion}

@@

The design choices and the results presented here are still preliminary.
While the general scheme we propose might be difficult to implement in
a general sense, we believe that it is a reasonable challenge to
implement it in the case of a personal robot assistant essentially
devoted to fetch-and-carry, as well as for interactive manipulation
tasks and associated activities.

\section*{Acknowledgment}

This work has been supported by EU FP7 ``SAPHARI'' under grant agreement no. ICT-287513.

\bibliographystyle{elsarticle-num}
\bibliography{laas_hri}
